---
title: "GOV51 Problem Set 4"
author: "Brian Jeon"
date: "04/17/2026"
format: pdf
header-includes:
  - \usepackage{changepage}
  - \usepackage{lmodern}
  - \usepackage{anyfontsize}
  - \fontsize{20pt}{20pt}\selectfont
  - \usepackage{tabularray}
  - \UseTblrLibrary{siunitx}
  - \usepackage{siunitx}
---

# 1 Setup and Data Exploration

Github link \href{https://github.com/brianjeon0722/gov51-ps4}{here}.

## 1.1 Load packages and data

```{r, echo = F}

library(dplyr)
library(tidyverse)
library(vroom)
library(kableExtra)
library(ggplot2)
library(haven)
library(infer)
library(moderndive)
library(estimatr)
library(sandwich)
library(lmtest)
library(modelsummary)
library(car)
library(ivreg)
library(plm)
library(oiplot)

card <- read_csv("data/card1995.csv")

```

## 1.2 Describe the data

3,010 observations. Each observation is an \textbf{individual}.

## 1.3 Distribution of years of education

$\bar{educ}$ = 13.26 years. Distribution is left skewed with a large spike at 12 (finishing high school) and 16 (finishing college). This makes sense because most people elect to finish their degrees rather than dropout. 

```{r}

card %>%
  ggplot(mapping = aes(x = educ)) + 
  geom_histogram() + 
  theme_oi() + 
  labs(x = "Education Years",
       y = "Count")

mean(card$educ, na.rm = TRUE)

card %>%
  group_by(educ) %>%
  summarize(n = n())

```

## 1.4 The naive relationship

Yes, based on the positive slope on the OLS regression, this plot suggests more education is associated with higher wages. 

```{r}

card %>%
  ggplot(mapping = aes(x = educ, y = lwage)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  theme_oi() +
  labs(x = "Education Years",
       y = "Log Wages")

```

## 1.5 Why OLS might be misleading.

Omitted Variable Bias: (1) it's possible that those who choose more years of education have some personality characteristic such as being more hardworking, intelligent, curious, etc. which is the actual cause of higher wages. (2) it's possible that those with more education years tend to come from wealthier families or have special backgrounds / connections that allow them to obtain higher wages. 

\newpage

# The Instrument: College Proximity

## 2.1 What the instrument measures

68.21\% of respondents grew up near a 4-year college at age 14. 

```{r}

mean(card$nearc4)

```

## 2.2 Relevance: does proximity affect education?

$nearc4$ estimates:
\begin{itemize}
\item $\hat{\alpha}_1 = 0.32711$
\item $SE(\hat{\alpha}_1) = 0.080554$
\item $t = 4.0607$
\end{itemize}

Holding race, living in metropolitan areas, living in the south, experience, and marriage constant, on average, those living near a 4-year college at age 14 have 0.32711 more education years than those who did not live near a 4-year college at age 14. This sign makes sense because living closer to a college gives you lower tuition (since your taxes go toward funding it) and lower moving / switching costs due to proximity, making it easier to attend college and thus increase your education years. 

```{r}

first_stage <- lm_robust(educ ~ nearc4 + black + smsa + south + exper + expersq 
                         + married, data = card, se_type = "HC2")

summary(first_stage)

```

## 2.3 Checking instrument strength

$F = 16.49 > 10$. The instrument is strong. Instrument strength matters because IV is biased in finite samples but consistent as $n \rightarrow \infty$. As a result, because IV tries to isolate the variation in endogenous variable $X_i$ that is random, if the instrument is weak, there is too little random variation for the 2SLS and this high variance leas to unreliable inferences. In other words, because Z (instrument) is used to predict X, then $\hat{X}$ is used to predict Y, if Z cannot predict X well, $\hat{X}$ is a poor estimate of true random variation in X and might still be endogenous. 

```{r}

4.0607^2

```

## 2.4 Exclusion restriction

a. Living near a 4-year college when young has \textbf{zero} effect on future log wages except through increasing education years. In other words, once we control for education, there shoul be no relationship between $nearc4$ and $lwage$.

b. Colleges often attract firms to the city to capitalize off the greater skilled labor, and many get jobs near the college they attend. As a result, there may be more higher paying roles for those living near colleges. Ex: Austin had Dell, Tesla, Texas Instruments, Apple, etc. move in due to the University of Texas. Therefore, being near a 4-year college may improve log wages not due to increases in education years but due to proximity to high paying jobs. 

## 2.5 Balance check

The groups are not balanced (see table below). These differences are statistically significant at the $\alpha = 0.05$ level (p-values for each is 0...) (see regression tables below). This imbalanace tells us that the instrument may be weak and cannot isolate the as-if random variation in education years. 

```{r}

card %>%
  group_by(nearc4) %>%
  summarize(mean_black = mean(black),
            mean_smsa = mean(smsa),
            mean_south = mean(south))

```

```{r}

summary(lm(black ~ nearc4, data = card))

summary(lm(smsa ~ nearc4, data = card))

summary(lm(south ~ nearc4, data = card))

```

\newpage

# 3 The Wald Estimator

## 3.1 Compute the reduced form

$\hat{\pi}_1 = 0.0408540$. Holding race, living in metropolitan areas, living in the south, experience, and marriage constant, on average, those living near a 4-year college at age 14 have 0.0408540 greater log wages than those who did not live near a 4-year college at age 14 ($\approx$ 4.09\% greater wages). This tells us that greater proximity to a college has an associated increase in wages. 

```{r}

summary(lm(lwage ~ nearc4 + black + smsa + south + exper + expersq 
                         + married, data = card))

```

## 3.2 Calculate the Wald estimator by hand

$\hat{\tau}_{Wald} = \frac{0.0408540}{0.32711} = 0.1248938$

## 3.3 Interpret the Wald estimator 

Holding race, living in metropolitan areas, living in the south, experience, and marriage constant, on average, the model predicts a 0.1249 increase in log wages for every 1 year increase in education years ($\approx$ 12.49\% increase in wages for every 1 year increase in education years)

\newpage

# 4 OLS vs. 2SLS

## 4.1 OLS estimate

$\hat{\beta}_1 = 0.071498$. Holding race, living in metropolitan areas, living in the south, experience, and marriage constant, on average, the model predicts a 0.071498 increase in log wages for every 1 year increase in education years ($\approx$ 7.1498\% increase in wages for every 1 year increase in education years)

```{r}

ols <- lm_robust(lwage ~ educ + black + smsa + south + exper + expersq + married,
                 data = card, se_type = "HC2")

summary(ols)

```

## 4.2 2SLS estimate

$educ$ coefficient = 0.124893. Yes, this matches the Wald estimate from \textbf{Section 3.2}. 

```{r}

iv <- iv_robust(lwage ~ educ + black + smsa + south + exper + expersq + married |
                  nearc4 + black + smsa + south + exper + expersq + married,
                data = card, se_type = "HC2")

summary(iv)

```

## 4.3 Side-by-Side comparison

```{r}

modelsummary(list("OLS" = ols, "2SLS" = iv), 
             coef_map = c("educ" = "Years of Education"),
             gof_map = c("nobs", "r.squared"))

```

## 4.4. Why do OLS and 2SLS differ?

a. 2SLS is larger.
b. (1) Compliers vs. Everyone: OLS estimates the effect of education on EVERYONE in the data. On the other hand, IV only looks at compliers (those who gain education years when they are proximate to a college). As a result, education may have higher returns for compliers compared to everyone (for those that greater education depends on proximity are likely influenced by distance, cost, etc. and they may benefit more from education). (2) Measurement Error: it's possible for education to be measured imprecisely since not all school records may be precise or may get deleted when a student drops out (measurement error on the X variable) which causes attentuation / downward bias. 
c. For the causal effect, IV is more credible if our 3 assumptions hold (relevance, exclusion, independence) because we may be concerned about OVB, reverse causality, or general endogenity ($\epsilon$ / error being correlated with X). IV isolates the as-if random variation in X and uses that random variation to predict Y which in theory, isolates the causal effect like a randomized experiment does. 

## 4.5 Adding IQ as a control

a. $educ$ coefficient = 0.071498 $\rightarrow$ 0.0675687. Including IQ decreased the coefficient. 

```{r}

summary(lm(lwage ~ educ + black + smsa + south + exper + expersq 
                         + married + IQ, data = card))

```

b. $educ$ coefficient = 0.124893 $\rightarrow$ 0.1093899. The coefficient decreases by 12.41\%. 

```{r}

# check

summary(iv_robust(lwage ~ educ + black + smsa + south + exper + expersq 
                  + married + IQ| nearc4 + black + smsa + south + exper + 
                    expersq + married + IQ,
                data = card, se_type = "HC2"))

```

c. Although both coefficients decline, implying that original OLS was not capturing the OVB of IQ (higher edu positively correlated with IQ and higher IQ posiively correlated with greater wages), the end estimate between OLS and IV are still different to each other (difference of 0.0418212). As a result, IQ does not explain the source of disrepancy between OLS and IV (likely measurement error and LATE vs. everyone). 

# 5 Interpretation and LATE

## 5.1 Who are the compliers?

The compliers are those who increase years of education when they are close to a 4-year college at age 14. In other words, without being close to a 4-year college at age 14, they would have lower years of education. These individuals are likely those who are sensitive to costs, distance, and proximity. As discussed in \textbf{2.2 Relevance: does proximity affect education?}, living close to a college reduces transportation, emotional, or financial costs and perhaps even tuition since your local tax dollars goes toward paying for the college and therefore local residents often get discounts. As a result, compliers are those where these factors are relevant in changing decision making (perhaps credit constrained, lower SES). 

## 5.2 Write the LATE interpretation

Holding race, living in metropolitan areas, living in the south, experience, and marriage constant, on average, the 2SLS estimate implies that log wages increase by 0.1248938 for every 1 year increase in education ($\approx$ 12.48938\% increase in wages for every 1 year increase in education) for those whose years of education increases by living close to a 4-year college at age 14. This effect was identified using variation in proximity to a 4-year college at age 14. 

## 5.3 External validity

No, this is likely not the same as the ATE. Those who are sensitive to distance or costs are likely very different from those who are not. Those who do not care about distance or transportation costs may have higher SES or have certain personalities that predispose them to higher wages meaning the marginal effect of education on higher wages may be limited (they could've gotten a good wage without the extra education). On the other hand, because compliers who are sensitive may already start with lower wages or have greater financial burden, the marginal effect of education on wages is likely higher compared to the typical person. 
